msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-07 paper.tex
Response to review.

Kingman's coalescent assumes that at most two ancestral lineages can merge at
each merger event. Although this is generally a reasonable approximation there
are certain situations in which the underlying mathematical assumptions are
violated. For example in certain highly fecund organisms
\citep{hedgecock_94,B94,HP11,A04,irwin16}, where individuals have the ability
to produce numbers of offspring on the order of the population size and
therefore a few individuals may produce the bulk of the offspring in any given
generation\citep{hedgecock_94}. These population dynamics violate basic
assumptions of the Kingman coalescent, and are better modelled by
`multiple-merger' coalescents \citep{DK99,P99,S99,S00,MS01}, in which more than
two lineages can merge in a given event. Multiple-merger coalescent processes
have also been shown to be relevant for modeling the effects of selection on
gene genealogies \citep{Gillespie909,DS04}.

Although multiple merger coalescents have been of significant theoretical
interest for around two decades, there has been little practical software
available to simulate these models.
\cite{kelleher2013coalescent,kelleher2014coalescent} developed packages to
simulate a related spatial continuum model~\citep{barton2010new},
and \cite{zhu2015hybrid} simulate genealogies within a species tree
based on a multiple-merger model.
More recently \cite{becheler2020occupancy}
provide a general method for simulating multiple merger processes
as part of the Quetzal framework~\citep{becheler2019quetzal}.
None of these methods work with large genomes, and very little work
has been performed on simulating multiple merger processes with recombination.

We have added two multiple merger coalescent models in \msprime\ 1.0, the
$\beta$-coalescent and ``dirac''-coalescent, allowing us to simulate
such models with recombination efficiently for the first time.
Please see the Appendix for full details and derivations.

@@

A larger class of multiple-merger coalescents involving simultaneous
multiple mergers of distinct groups of ancestral lineages also exists
\citep{S00}. These are commonly referred to as $\Xi$-coalescents, and
can be shown to be the limits of ancestral processes derived from
population models incorporating diploidy (or more general polyploidy)
\citep{BBE13,Blath2016}, and certain models of selection \citep{DS04}.
To describe a general $\Xi$-coalescent let $\Delta$ denote the
infinite simplex \be\label{Delta} \Delta := \{ (x_1, \ldots ): x_1 \ge
x_2 \ge \cdots \ge 0, \sum_{j}x_j \le 1\}; \ee for any
$r \in \{1,2, \ldots\}$ let $k_1, \cdots, k_r \ge 2$, and
$b = s + k_1 + \cdots + k_r$ be the total number of blocks in a given
partition ($s = b - k_1 - \cdots - k_r$ is the number of blocks not
participating in the mergers at the given time).  The existence of
simultaneous multiple-merger coalescents was proved by \cite{S00}.
There exists a finite measure $\Xi$ on $\Delta$, with
$\Xi = \Xi_0 + a\delta_0$, the measure $\Xi_0$ has no atom at zero,
$a >0$ is fixed, and the rate at which one sees a given (simultaneous)
merger of ancestral lineages with merger sizes $k_1, \ldots, k_r$,
$s = n - k_1 - \cdots - k_r$, is given by
\begin{esplit}{xi}
  \lambda_{n; k_1, \ldots, k_r; s}  & = \int_\Delta  \sum_{\ell = 0}^s \sum_{\substack {i_1, \ldots, i_{r+\ell} = 1\\ \text{all distinct}} }^\infty  \binom{s}{\ell} x_{i_1}^{k_1} \cdots  x
_{i_{r}}^{k_r} x_{i_{r+1}} \cdots x_{i_{r+\ell}}\left(1 - \sum_{j=1}^\infty x_j \right)^{s-\ell} \frac{1}{ \sum_{j=1}^\infty x_j^2 } \Xi_0(dx)   \\
  & +  a\one{r=1, k_1 = 2}.
\end{esplit}%
The number of such $(k_1, \ldots, k_r)$ mergers is
\be\label{N}
    \mathcal{N}(b; k_1, \ldots, k_r ) = \binom{b}{k_1 \ldots k_r\, s} \frac{1}{ \prod_{j=2}^b\ell_j!  },
\ee
\citep{S00},   in particular  $\mathcal{N}(b;2) = b(b-1)/2$, and one can compute the total rate of a  $(k_1, \ldots, k_r)$ merger as
\be\label{lambdabkall}
      \lambda(n; k_1, \ldots, k_r)         =    \mathcal{N}(b; k_1, \ldots, k_r ) \lambda_{n; k_1, \ldots, k_r; s}.
\ee

The total rate is given by, with
$n\ge 2$ denoting the total  number of ancestral lineages,
\be
\label{Xilambdab}
\lambda_{n} = \int_\Delta \left(1 - \sum_{\ell = 0}^n \sum_{i_1 \neq
\cdots \neq i_\ell } \binom{n}{\ell} x_{i_1}\cdots x_{i_\ell}\left(1
- \sum_{i=1}^\infty x_i\right)^{n-\ell} \right)
\frac{1}{\sum_{j=1}^\infty x_j^2}\Xi_0(dx) + a\binom{n}{2} \ee
\citep{S00}.  Viewing coalescent processes strictly as mathematical
objects, it is clear that the class of $\Xi$-coalescents contains
$\Lambda$-coalescents as a specific example (i.e.\ allowing at most
one group of lineages to merge each time), and the class of
$\Lambda$-coalescents contain the Kingman-coalescent as a special
case.  However, viewed as limits of ancestral processes derived from
specific population models they are not nested, since one would
obtain $\Lambda$-coalescents when deriving coalescent processes from
haploid population models incorporating sweepstakes reproduction and
high fecundity, and $\Xi$-coalescents for diploid populations.  One
should therefore apply the models as appropriate, i.e.\
$\Lambda$-coalescents to data (e.g.\ mtDNA data) inherited in a
haploid fashion, and $\Xi$-coalescents to e.g.\ autosomal data
inherited in diploid or polyploid fashion \citep{Blath2016}.

@@

The other multiple-merger coalescent model incorporated in \msprime\ is derived
from an adaptation of the haploid population model considered by
\cite{schweinsberg03} to diploid populations \citep{BLS15}.  In the haploid
version, in each generation individuals independently produce random numbers of
juveniles, where the random  number of juveniles $(X)$ produced by a given
individual  has the stable law \be\label{jX} \lim_{k\to \infty} C k^\alpha
\prb{X \ge k} = 1 \ee with index $\alpha > 0$, and $C > 0$ is a normalising
constant.   One can interpret  Eq.~\eqref{jX} as  stating the form of the
probability distribution  for number of juveniles at least on the order of the
population size.     If the random  total number of juveniles $(S_N)$ produced
in this way is at least the population size $(N)$, then one samples  $N$
juveniles uniformly at random without replacement to form the next generation
of  reproducing individuals; if  $S_N < N$ one simply  carries on with  the
same  set of individuals.   However,   assuming  $\EE{X} > 1$, one can show
that $\prb{S_N < N}$ decays exponentially fast in $N$ \citep{schweinsberg03}.
If $\alpha \ge 2$ the ancestral process converges to the Kingman-coalescent; if
$1 \le \alpha < 2$ the ancestral process converges to a specific case of a
$\Lambda$-coalescent, where the $\Lambda$ measure in Eq.~\eqref{lambdabk} is
associated with the Beta$(2-\alpha, \alpha)$ probability distribution, i.e.\
\be\label{Fbeta} \Lambda(dx) = \one{0< x \le 1} \frac{1}{B(2-\alpha,\alpha)}
x^{1 - \alpha}(1-x)^{\alpha - 1}  dx, \ee where $B(2-\alpha,\alpha)$ is the
beta function $B(a,b) = \Gamma(a)\Gamma(b)/\Gamma(a+b)$, $a,b > 0$
\citep{schweinsberg03}.
This model has been adapted to diploid populations
by \cite{BLS15}, where  the resulting coalescent process
is a  four-fold $\Xi$-coalescent on  $(x/4, x/4, x/4, x/4, 0, 0, \ldots)$,
where $x$ is a random variate with  the Beta$(2-\alpha,\alpha)$ distribution.
The merger rate (see Eq.~\eqref{xi}) is then ($1 \le r \le 4$)
\be\label{xibeta} \lambda_{b;k_1, \ldots, k_r} = \sum_{\ell = 0}^{ (b -
k)\wedge (4-r) } \binom{b-k}{\ell} \frac{ (4)_{r+\ell} }{4^{k+\ell}}
\frac{B(k+\ell - \alpha, b-k-\ell + \alpha ) }{B(2-\alpha,\alpha)} \ee
\citep{Blath2016,BLS15}. The interpretation of Eq.~\eqref{xibeta} is  that the
random   number of diploid  juveniles each  diploid pair of parents  produces
is  governed by  the law in Eq.~\eqref{jX},   and   each diploid juvenile  is
assigned  one chromosome  from each parent (selfing is excluded).

>>>>

Kingman's coalescent assumes that only two ancestral lineages can merge at
each merger event. Although this is generally a reasonable approximation there
are certain situations in which the underlying mathematical assumptions are
violated. For example in certain highly fecund organisms
\citep{hedgecock_94,B94,HP11,A04,irwin16}, where individuals have the ability
to produce numbers of offspring on the order of the population size and
therefore a few individuals may produce the bulk of the offspring in any given
generation\citep{hedgecock_94}. These population dynamics violate basic
assumptions of the Kingman coalescent, and are better modelled by
`multiple-merger' coalescents \citep{DK99,P99,S99,S00,MS01}, in which more than
two lineages can merge in a given event. Multiple-merger coalescent processes
have also been shown to be relevant for modeling the effects of selection on
gene genealogies \citep{Gillespie909,DS04}.

Although multiple merger coalescents have been of significant theoretical
interest for around two decades, there has been little practical software
available to simulate these models.
\cite{kelleher2013coalescent,kelleher2014coalescent} developed packages to
simulate a related spatial continuum model~\citep{barton2010new},
\cite{zhu2015hybrid} simulate genealogies within a species tree
based on a multiple-merger model, and
\cite{becheler2020occupancy} provide a general method for simulating
multiple merger processes
as part of the Quetzal framework~\citep{becheler2019quetzal}.
The \texttt{Beta-Xi-Sim} simulator~\citep{koskela2018multi,koskela2019robust}
includes a number of extensions to the Beta-coalescent, although
it only supports unlinked loci.
None of these methods work with large genomes, and very little work
has been performed on simulating multiple merger processes with recombination.

We have added two multiple merger coalescent models in \msprime\ 1.0, the
Beta-coalescent and ``dirac''-coalescent, allowing us to simulate
such models with recombination efficiently for the first time.
These simulation models have been extensively validated against
analytical results from the site frequency
spectrum~\citep{birkner2013statistical,blath2016site,hobolth2019phase}
 as well as more general properties of coalescent processes.
Please see the Appendix for more details and model derivations.

@@

A larger class of multiple-merger coalescents involving simultaneous
multiple mergers of distinct groups of ancestral lineages also exists
\citep{S00}. These are commonly referred to as $\Xi$-coalescents, and
can be shown to be the limits of ancestral processes derived from
population models incorporating diploidy (or more general polyploidy)
\citep{BBE13,blath2016site}, and certain models of selection \citep{DS04}.
To describe a general $\Xi$-coalescent let $\Delta$ denote the
infinite simplex \be\label{Delta} \Delta := \{ (x_1, \ldots ): x_1 \ge
x_2 \ge \cdots \ge 0, \sum_{j}x_j \le 1\}; \ee for any
$r \in \{1,2, \ldots\}$ let $k_1, \cdots, k_r \ge 2$, and
$b = s + k_1 + \cdots + k_r$ be the total number of blocks in a given
partition ($s = b - k_1 - \cdots - k_r$ is the number of blocks not
participating in the mergers at the given time).  The existence of
simultaneous multiple-merger coalescents was proved by \cite{S00}.
There exists a finite measure $\Xi$ on $\Delta$, with
$\Xi = \Xi_0 + a\delta_0$, the measure $\Xi_0$ has no atom at zero,
$a >0$ is fixed, and the rate at which one sees a given (simultaneous)
merger of ancestral lineages with merger sizes $k_1, \ldots, k_r$,
$s = n - k_1 - \cdots - k_r$, is given by
\begin{esplit}{xi}
  \lambda_{n; k_1, \ldots, k_r; s}  & = \int_\Delta  \sum_{\ell = 0}^s \sum_{\substack {i_1, \ldots, i_{r+\ell} = 1\\ \text{all distinct}} }^\infty  \binom{s}{\ell} x_{i_1}^{k_1} \cdots  x
_{i_{r}}^{k_r} x_{i_{r+1}} \cdots x_{i_{r+\ell}}\left(1 - \sum_{j=1}^\infty x_j \right)^{s-\ell} \frac{1}{ \sum_{j=1}^\infty x_j^2 } \Xi_0(dx)   \\
  & +  a\one{r=1, k_1 = 2}.
\end{esplit}%
The number of such $(k_1, \ldots, k_r)$ mergers is
\be\label{N}
    \mathcal{N}(b; k_1, \ldots, k_r ) = \binom{b}{k_1 \ldots k_r\, s} \frac{1}{ \prod_{j=2}^b\ell_j!  },
\ee
\citep{S00},   in particular  $\mathcal{N}(b;2) = b(b-1)/2$, and one can compute the total rate of a  $(k_1, \ldots, k_r)$ merger as
\be\label{lambdabkall}
      \lambda(n; k_1, \ldots, k_r)         =    \mathcal{N}(b; k_1, \ldots, k_r ) \lambda_{n; k_1, \ldots, k_r; s}.
\ee

The total rate is given by, with
$n\ge 2$ denoting the total  number of ancestral lineages,
\be
\label{Xilambdab}
\lambda_{n} = \int_\Delta \left(1 - \sum_{\ell = 0}^n \sum_{i_1 \neq
\cdots \neq i_\ell } \binom{n}{\ell} x_{i_1}\cdots x_{i_\ell}\left(1
- \sum_{i=1}^\infty x_i\right)^{n-\ell} \right)
\frac{1}{\sum_{j=1}^\infty x_j^2}\Xi_0(dx) + a\binom{n}{2} \ee
\citep{S00}.  Viewing coalescent processes strictly as mathematical
objects, it is clear that the class of $\Xi$-coalescents contains
$\Lambda$-coalescents as a specific example (i.e.\ allowing at most
one group of lineages to merge each time), and the class of
$\Lambda$-coalescents contain the Kingman-coalescent as a special
case.  However, viewed as limits of ancestral processes derived from
specific population models they are not nested, since one would
obtain $\Lambda$-coalescents when deriving coalescent processes from
haploid population models incorporating sweepstakes reproduction and
high fecundity, and $\Xi$-coalescents for diploid populations.  One
should therefore apply the models as appropriate, i.e.\
$\Lambda$-coalescents to data (e.g.\ mtDNA data) inherited in a
haploid fashion, and $\Xi$-coalescents to e.g.\ autosomal data
inherited in diploid or polyploid fashion \citep{blath2016site}.

@@

The other multiple-merger coalescent model incorporated in \msprime\ is derived
from an adaptation of the haploid population model considered by
\cite{schweinsberg03} to diploid populations \citep{BLS15}.  In the haploid
version, in each generation individuals independently produce random numbers of
juveniles, where the random  number of juveniles $(X)$ produced by a given
individual  has the stable law \be\label{jX} \lim_{k\to \infty} C k^\alpha
\prb{X \ge k} = 1 \ee with index $\alpha > 0$, and $C > 0$ is a normalising
constant.   One can interpret  Eq.~\eqref{jX} as  stating the form of the
probability distribution  for number of juveniles at least on the order of the
population size.     If the random  total number of juveniles $(S_N)$ produced
in this way is at least the population size $(N)$, then one samples  $N$
juveniles uniformly at random without replacement to form the next generation
of  reproducing individuals; if  $S_N < N$ one simply  carries on with  the
same  set of individuals.   However,   assuming  $\EE{X} > 1$, one can show
that $\prb{S_N < N}$ decays exponentially fast in $N$ \citep{schweinsberg03}.
If $\alpha \ge 2$ the ancestral process converges to the Kingman-coalescent; if
$1 \le \alpha < 2$ the ancestral process converges to a specific case of a
$\Lambda$-coalescent, where the $\Lambda$ measure in Eq.~\eqref{lambdabk} is
associated with the Beta$(2-\alpha, \alpha)$ probability distribution, i.e.\
\be\label{Fbeta} \Lambda(dx) = \one{0< x \le 1} \frac{1}{B(2-\alpha,\alpha)}
x^{1 - \alpha}(1-x)^{\alpha - 1}  dx, \ee where $B(2-\alpha,\alpha)$ is the
beta function $B(a,b) = \Gamma(a)\Gamma(b)/\Gamma(a+b)$, $a,b > 0$
\citep{schweinsberg03}.
This model has been adapted to diploid populations
by \cite{BLS15}, where  the resulting coalescent process
is a  four-fold $\Xi$-coalescent on  $(x/4, x/4, x/4, x/4, 0, 0, \ldots)$,
where $x$ is a random variate with  the Beta$(2-\alpha,\alpha)$ distribution.
The merger rate (see Eq.~\eqref{xi}) is then ($1 \le r \le 4$)
\be\label{xibeta} \lambda_{b;k_1, \ldots, k_r} = \sum_{\ell = 0}^{ (b -
k)\wedge (4-r) } \binom{b-k}{\ell} \frac{ (4)_{r+\ell} }{4^{k+\ell}}
\frac{B(k+\ell - \alpha, b-k-\ell + \alpha ) }{B(2-\alpha,\alpha)} \ee
\citep{blath2016site,BLS15}. The interpretation of Eq.~\eqref{xibeta} is  that the
random   number of diploid  juveniles each  diploid pair of parents  produces
is  governed by  the law in Eq.~\eqref{jX},   and   each diploid juvenile  is
assigned  one chromosome  from each parent (selfing is excluded).
